## The dataset is downloaded from kaggle.
## This is implemented in Posit Studio cloud.
## Below is the source code written to develop an R markdown file.
## This is the YAML header section
---
title: "Hotel bookings"
author: "Uzma Anjum"
date: "2024-05-23"
output: html_document
---

## Setting up environment
Initially I need to set up my R environment by installing and loading required packages such as 'tidyverse','skimr','janitor','dplyr' and 'ggplot2'packages.
```{r}
install.packages("tidyverse")
library(tidyverse)
```


## Import dataset with a new dataframe
```{r}
bookings_df <- read_csv("hotel_bookings.csv")
```

## Get an overview of data



```{r}
glimpse(bookings_df)

```

```{r}
colnames(bookings_df)
```

## Manipulate data 
Here I am arranging data in ascending and descending orders on specific columns to focus on bookings that were made in advance
```{r}
arrange(bookings_df, lead_time)
arrange(bookings_df, desc(lead_time))
```

## New dataframe bookings_df_v2
And then i created a new dataframe to save the changes made by arrange function
```{r}
bookings_df_v2 <- arrange(bookings_df, desc(lead_time))
```

## Find out max and min values
finding out the maximum and minimum lead times without sorting the whole dataset using the `arrange()` function. Try it out using the max() and min() functions below:
```{r}
max(bookings_df$lead_time)
min(bookings_df$lead_time)
mean(bookings_df$lead_time)
```

## Filter data
First step will be creating a new data set that only contains data about city hotels.It is done by using the `filter()` function, and name new data frame 'bookings_city':
```{r}
bookings_city <-  filter(bookings_df, bookings_df$hotel=="City Hotel")
```
```{r}
head(bookings_city)
```

### Check what the average lead time for this set of hotels is, just like for all of hotels before:
```{r}
mean(bookings_city$lead_time)
```

### Using 'groupby()' and 'summarize()'
Now, lets assume I want to know a lot more information about city hotels, including the maximum and minimum lead time.Am also interested in how they are different from resort hotels. I don't want to run each line of code over and over again, so I decide to use the `group_by()`and`summarize()` functions.I can also use the pipe operator to make my code easier. I will store the new data set in a data frame named 'hotel_summary':

```{r}
bookings_summary <- bookings_df %>%
                    group_by(hotel) %>%
                    summarise(average_lead_time=mean(lead_time),
                    min_lead_time=min(lead_time),
                    max_lead_time=max(lead_time))
```
Now check it out using head() function.
```{r}
head(bookings_summary)
```

## Cleaning data
Previously I was primarily interested in the following variables: hotel, is_canceled, lead_time. Creating a new data frame with just those columns, calling it `trimmed_df`.
```{r}
trimmed_df <- bookings_df %>%  
              select(hotel, is_canceled, lead_time) 
```

## Rename() function 
Renaming hotel to hotel_type for clarity
```{r}
trimmed_df %>%  
select(hotel, is_canceled, lead_time) %>% 
rename(hotel_type = hotel)
```

## Unite() function
Combine arrival month and year using unite function and store it in new data frame 'Arrival_df'
```{r}
arrival_df <- bookings_df %>%   
select(arrival_date_year, arrival_date_month) %>% 
unite(arrival_month_year, c("arrival_date_month", "arrival_date_year"), sep = " ")
```

## Mutate() function 
Create new column to calculate total guest including adults, children,babies
```{r}
total_guests_df <- bookings_df %>%  
  mutate(guests = adults + children + babies)
```

## Summary Statistics
Now, calculating summary statistics such as total number of canceled bookings, avg lead time for bookings
```{r}
statistics_df <- bookings_df %>%                 
    summarize(number_canceled = sum(is_canceled),
    average_lead_time = mean(lead_time))
```

```{r}
head(statistics_df)
```

# Data visualization
To determine if people with children book hotel rooms in advance.

```{r warning=FALSE}
ggplot(data = bookings_df) +               
  geom_point(mapping = aes(x = lead_time, y = children))

```

To map, place 'stays_in_weekend_nights' on the x-axis and 'children' on the y-axis 
```{r}
ggplot(data =bookings_df) +                                           
  geom_point(mapping = aes(x = stays_in_weekend_nights, y = children))
```

Previously, `geom_point` was used to make a scatter plot comparing lead time and number of children. Now,`geom_bar` is used to make a bar chart in this code chunk: 
```{r}
ggplot(data = bookings_df) +
  geom_bar(mapping = aes(x = distribution_channel))
```

Now,if I want to know the number of bookings for each distribution type is different depending on whether or not there was a deposit or what market segment they represent. 
Am using'fill=deposit_type' to accomplish this.
```{r}
ggplot(data = bookings_df) +
  geom_bar(mapping = aes(x = distribution_channel, fill=deposit_type))
```

Now below is the code to answer the question about different market segments.We will use 'fill=market_segment' to accomplish this.
```{r}
ggplot(data = bookings_df) +
  geom_bar(mapping = aes(x = distribution_channel, fill=market_segment))

```

After reviewing the charts let us create separate charts for each deposit type and market segment to help them understand the differences more clearly.

## Chart for each deposit type
```{r}
ggplot(data = bookings_df) +
  geom_bar(mapping = aes(x = distribution_channel)) +
  facet_wrap(~deposit_type)
```

## Chart for each market segment
```{r}
ggplot(data = bookings_df) +
  geom_bar(mapping = aes(x = distribution_channel)) +
  facet_wrap(~market_segment)
```

## 'facet grid' function
The `facet_grid` function does something similar. The main difference is that `facet_grid` will include plots even if they are empty. Run the code chunk below to check it out:
```{r}
ggplot(data = bookings_df) +
  geom_bar(mapping = aes(x = distribution_channel)) +
  facet_grid(~deposit_type)
```

## All in one chart

Now, lets put all of this in one chart and explore the differences by deposit type and market segment.

Notice how the ~ character is being used before the variables that the chart is being split by:
```{r}
ggplot(data = bookings_df) +
  geom_bar(mapping = aes(x = distribution_channel)) +
  facet_wrap(~deposit_type~market_segment)
```

## creating a bar chart
Showing each hotel type and market segment.Use different colors to represent each market segment:
```{r}
ggplot(data = bookings_df) +
  geom_bar(mapping = aes(x = hotel, fill = market_segment))
```

Lets use the facet_wrap() function to create a separate plot for each market segment:
```{r}
ggplot(data = bookings_df) +
  geom_bar(mapping = aes(x = hotel)) +
  facet_wrap(~market_segment)
```

## Add title
The first thing people will pay attention to when they encounter a data visualization for the first time is at tilte. To add a title, we will add `labs()` at the end of `ggplot()` command and then input a title there as shown below:
```{r}
ggplot(data = bookings_df) +
  geom_bar(mapping = aes(x = market_segment)) +
  facet_wrap(~hotel) +
  labs(title="Comparison of market segments by hotel type for hotel bookings")
```

Also  adding another detail about what time period this data covers. To do this, find out when the data is from. 

To do this I used `min()` and 'max()'function on the year column in the data
```{r}
min(bookings_df$arrival_date_year)
```

```{r}
max(bookings_df$arrival_date_year)
```

##Save them

Here i need to save them as varaibales in order to use them in labelling.
```{r}
mindate <- min(bookings_df$arrival_date_year)
```

```{r}
maxdate <- max(bookings_df$arrival_date_year)
```

## Subtitles

lets add subtitle in the labs() function
```{r}
ggplot(data = bookings_df) +
  geom_bar(mapping = aes(x = market_segment)) + facet_wrap(~hotel) +
  labs(title="Comparison of market segments by hotel type for hotel bookings",
       subtitle=paste0("Data from: ", mindate, " to ", maxdate))
```

## Caption

Switching subtitle to caption which will be in bottom right corner
```{r}
ggplot(data = bookings_df) +
  geom_bar(mapping = aes(x = market_segment)) + facet_wrap(~hotel) +
  labs(title="Comparison of market segments by hotel type for hotel bookings",
       caption=paste0("Data from: ", mindate, " to ", maxdate))
```

## Making x-axis and y-axis labels very clear

```{r}
ggplot(data = bookings_df) +
  geom_bar(mapping = aes(x = market_segment)) + facet_wrap(~hotel) +
  labs(title="Comparison of market segments by hotel type for hotel bookings",
       caption=paste0("Data from: ", mindate, " to ", maxdate),
       x="Market Segment",
       y="Number of Bookings")
```

## Saving chart
```{r}
ggsave('bookings_df.png',
       width=16,
       height=8)

```


